Tissue Enrichment Analysis for C. elegans Genomics

### 2016-04-30 tests/hgf_benchmarking/doc/tea_main_text.tex
writing
stupid bmc

\begin{backmatter}

\section*{Competing interests}
 	The authors declare that they have no competing interests.
\section*{Author's contributions}
    DA and PWS conceived of the project; DA developed algorithm;
	RYL made intellectual contributions to the project; RYL and JC
	developed the web GUI.
\section*{Acknowledgements}
	We thank Justin Bois for his help and support. 
	We would like to acknowledge all members of the Sternberg
	lab for helpful discussion.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Figures describing algorithm
\begin{figure}[h!]
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
   	 	\caption{}
    \label{fig:simdiagram}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
	      \caption{}
    \label{fig:trim_ends}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
    	\caption{}
    \label{fig:trim_roots}
  \end{subfigure}

@@

%gui
\begin{figure}
	\centering
    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
  	\captionsetup{width= 0.95\textwidth}
 	\caption{
	\csentence{Screenshot of results from the web GUI.} 
	After inputting a gene-list, the user is provided with the results. An HTML table is output with hyperlinks to the ontology terms. A publication-ready graph is provided below, which can be saved by  dragging to the desktop. The graph is colored for better visualization; color is not intended to convey information. The graph and the table show anatomy terms in  human-readable format, followed by their unique WBbt ID. Finally, lists  of the genes used and discarded for the analysis are also presented.
  }
  \label{fig:GUIresults}
\end{figure}

% workflow 
\begin{figure}
	\centering
	\includegraphics[width=0.95\textwidth]{figures/workflow.pdf}
	\captionsetup{width= 0.95\textwidth}
	\caption{
	\csentence{TEA Workflow.}
	The complete ontology is annotated continuously by WormBase curators. After each update, the ontology is processed to remove uninformative terms. The trimmed ontology is used for statistical testing. 
	Users can select a gene list and input it into our tool using our WormBase portal. The gene list is tested for enrichment using the trimmed ontology, and results are output in tabular and graphic formats for analysis. 
	}
	\label{fig:workflow}
\end{figure}

% statistical testing and validation Figures
%q plot
\begin{figure}[h!]
	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Kernel density estimates (KDE) for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.
	  }
	  \label{fig:qvals}
\end{figure}

%gaba
\begin{figure}
  \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
  \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
  \textbf{Set 2)} GABAergic gene set from Spencer~\cite{Spencer2011}.
  Arrowheads highlight identical terms between both analyses. All terms refer to neurons or neuronal tissues and are GABA-associated. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
   }
  \label{fig:intragree}
\end{figure}

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{
	\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools. 
	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
	\textbf{b)} TEA results, sorted by q-value (lowest on top) and fold-change. 
	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.
	}
	\label{fig:Dcon}
\end{figure}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}

\begin{table}[h!]
	\caption{
	Parameter specifications and number of tissues for all dictionaries. The `Method' column refers to the trimming criterion for the similarity metric. We used two such criteria, `any' and `avg'.`any': For a given sister set, if any sister had a similarity exceeding the corresponding threshold, all sisters were removed from the final dictionary. `avg': For a given sister set, if the average similarity across all the sisters in the set was greater than the corresponding threshold, all sisters were removed from the final dictionary.
	}
	\csvreader[tabular= c | c | c | c,
	table head= {\textbf{Annotation Cutoff}} & {\textbf{Similarity Threshold}} & {\textbf{Method}} & {\textbf{No. Of Terms in Dictionary}}\\\hline,
	late after line= \\,
	late after last line= \\\hline
	]
	{figures/TissueNumbers.csv}{}
	{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv}

	% \csvautotabular{figures/TissueNumbers.csv}
	\label{tab:DictionarySpecs}
\end{table}

\begin{table}[h!]
	\caption{
	Comparison of results for a neuronal-enriched gene set from Watson~\cite{Watson2008a}. We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'. In the table, columns are labeled with their significance value (Q-value) or enrichment fold change followed by a hyphen and a number which indicates which the cutoff for the dictionary that was used for testing. Not all tissues are present in either dictionary. Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries. Full table is available in Additional File 2.
	}

	\begin{adjustbox}{width=1\textwidth}
		\centering
		% \csvreader{figures/dict-comparison-50-33.csv}
		\csvreader[tabular={ r | c | c | c | c},
	    table head ={\textbf{Tissue}} & {\textbf{Q-value-33}} & {\textbf{Q-value-50}} & {\textbf{Enrichment-33}} & {\textbf{Enrichment-50}}\\\hline,
	    late after line= \\,
		late after last line=\\\hline
		]%
		{figures/dict-comparison-50-33.csv}{}	
		{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv}	
		\label{tab:interagree}
	\end{adjustbox}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- IPython Notebook}
    Tutorial for users interested in using our software within a python script
  \subsection*{Additional file 2 --- Folder Structure for SI files 3 and 4}
    A file detailing the folder structure of the zipped folders 3 and 4. 
  \subsection*{Additional file 3 --- Golden Gene Sets}
    A list of all the genes used for our benchmarking process. 
  \subsection*{Additional file 4 --- Results}
    A folder containing a complete version of the results we generated for this paper.

\end{backmatter}
\end{document}

>>>>

\begin{backmatter}
\section*{Abbreviations}

\begin{itemize}
	\item TEA --- Tissue Enrichment Analysis
	\item GO --- Gene Ontology
	\item WBbt ID --- A unique ID assigned to reference ontology terms
	\item WBgene ID -- A unique ID assigned to reference nematode genes
\end{itemize}

\section*{Ethics Approval and Consent to Participate}

Not applicable

\section*{Consent for Publication}

Not applicable

\section*{Competing interests}
 	The authors declare that they have no competing interests.
\section*{Author's contributions}
    DA and PWS conceived of the project; DA developed algorithm;
	RYL made intellectual contributions to the project; RYL and JC
	developed the web GUI.
\section*{Acknowledgements}
	We thank Justin Bois for his help and support. 
	We would like to acknowledge all members of the Sternberg
	lab for helpful discussion.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Additional Files}
  \subsection*{Additional file 1 --- TEA Tutorial}
    Tutorial for users interested in using our software within a python script
  \subsection*{Additional file 2 --- Folder Structure for SI files 3 and 4}
    A file detailing the folder structure of the zipped folders 3 and 4. 
  \subsection*{Additional file 3 --- Golden Gene Sets}
    A list of all the genes used for our benchmarking process. 
  \subsection*{Additional file 4 --- Results}
    A folder containing a complete version of the results we generated for this paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Figures describing algorithm
\begin{figure}[htbp]
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
   	 	\caption{}
    \label{fig:simdiagram}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
	      \caption{}
    \label{fig:trim_ends}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
    	\caption{}
    \label{fig:trim_roots}
  \end{subfigure}

@@

%gui
\begin{figure}[htbp]
	\centering
    % \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
  	\captionsetup{width= 0.95\textwidth}
 	\caption{
	\csentence{Screenshot of results from the web GUI.} 
	After inputting a gene-list, the user is provided with the results. An HTML table is output with hyperlinks to the ontology terms. A publication-ready graph is provided below, which can be saved by  dragging to the desktop. The graph is colored for better visualization; color is not intended to convey information. The graph and the table show anatomy terms in  human-readable format, followed by their unique WBbt ID. Finally, lists  of the genes used and discarded for the analysis are also presented.
  }
  \label{fig:GUIresults}
\end{figure}

% workflow 
\begin{figure}[htbp]
	\centering
	% \includegraphics[width=0.95\textwidth]{figures/workflow.pdf}
	\captionsetup{width= 0.95\textwidth}
	\caption{
	\csentence{TEA Workflow.}
	The complete ontology is annotated continuously by WormBase curators. After each update, the ontology is processed to remove uninformative terms. The trimmed ontology is used for statistical testing. 
	Users can select a gene list and input it into our tool using our WormBase portal. The gene list is tested for enrichment using the trimmed ontology, and results are output in tabular and graphic formats for analysis. 
	}
	\label{fig:workflow}
\end{figure}

% statistical testing and validation Figures
%q plot
\begin{figure}[htbp]
	% \includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Kernel density estimates (KDE) for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.
	  }
	  \label{fig:qvals}
\end{figure}

%gaba
\begin{figure}
  % \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
  \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
  \textbf{Set 2)} GABAergic gene set from Spencer~\cite{Spencer2011}.
  Arrowheads highlight identical terms between both analyses. All terms refer to neurons or neuronal tissues and are GABA-associated. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
   }
  \label{fig:intragree}
\end{figure}

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[htbp]
    % \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{
	\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools. 
	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
	\textbf{b)} TEA results, sorted by q-value (lowest on top) and fold-change. 
	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.
	}
	\label{fig:Dcon}
\end{figure}

@@

%% Use of \listoftables is discouraged.
%

\section*{Tables}

\begin{table}[p]
	\caption{
	Parameter specifications and number of tissues for all dictionaries. The `Method' column refers to the trimming criterion for the similarity metric. We used two such criteria, `any' and `avg'.`any': For a given sister set, if any sister had a similarity exceeding the corresponding threshold, all sisters were removed from the final dictionary. `avg': For a given sister set, if the average similarity across all the sisters in the set was greater than the corresponding threshold, all sisters were removed from the final dictionary.
	}
	% \csvreader[tabular= c | c | c | c,
	% table head= {\textbf{Annotation Cutoff}} & {\textbf{Similarity Threshold}} & {\textbf{Method}} & {\textbf{No. Of Terms in Dictionary}}\\\hline,
	% late after line= \\,
	% late after last line= \\\hline
	% ]
	% {figures/TissueNumbers.csv}{}
	% {\csvcoli & \csvcolii & \csvcoliii & \csvcoliv}

	\label{tab:DictionarySpecs}
\end{table}

\begin{table}[p]
	\caption{
	Comparison of results for a neuronal-enriched gene set from Watson~\cite{Watson2008a}. We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'. In the table, columns are labeled with their significance value (Q-value) or enrichment fold change followed by a hyphen and a number which indicates which the cutoff for the dictionary that was used for testing. Not all tissues are present in either dictionary. Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries. Full table is available in Additional File 2.
	}

	% \begin{adjustbox}{width=1\textwidth}
	% 	\centering
	% 	% \csvreader{figures/dict-comparison-50-33.csv}
	%
	% 	\csvreader[tabular={ r | c | c | c | c},
	% 		    table head ={\textbf{Tissue}} & {\textbf{Q-value-33}} & {\textbf{Q-value-50}} & {\textbf{Enrichment-33}} & {\textbf{Enrichment-50}}\\\hline,
	% 		    late after line= \\,
	% 	late after last line=\\\hline
	% 	]%
	% 	{figures/dict-comparison-50-33.csv}{}
	% 	{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv}
	%
	% \end{adjustbox}
	\label{tab:interagree}

\end{table}

\end{backmatter}

\end{document}
